|

### 2019-07-31 article/article.Rmd
update

---
title:  |
        | The Pitfalls of Bottom-Up Accountability: Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Guilherme Fasolin, Stephen Herzog, and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research and technical assistance. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "We study the effect of a mobile phone application designed to increase citizens' ability to monitor public works in Brazil. The app allowed individuals to find delayed school constructions in XXX Brazilian cities and send anonymous messages to politicians requestions information about construction delays and expected completion times. Our results show that the app has a null impact on the completion times subsequently reported by the municipalities. Additionally, we find that few politicians reacted to citizens' requests. These findings suggest it is difficult to motivate bottom-up accountability in new democracies, especially when politicians are unresponsive to non-electoral pressures. This paper has implications for the design of bottom-up accountability mechanisms."
abstractspacing: double
keywords: accountability, Brazil, field experiment, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

One of the key determinants of efficient public service provision is a robust accountability system \citep{besley2003incentives,cameron2004public,o1998horizontal}. In its standard definition, accountability is understood as the process of holding authorities into account for their actions \citep{finer1941administrative,mulgan2000accountability,o1990bureaucratic}. Past studies show that accountability ensures that politicians act on behalf of voters \citep{freire2010ngp, moncrieffe1998reconceptualizing}, reduces the opportunities for rent-seeking and corruption \citep{deininger2005does,wenar2006accountability} and increases the efficiency of the public sector \citep{adsera2003you,bjorkman2009power}. Moreover, greater accountability is associated with higher levels of economic growth, not only because it limits state discretion in the economy, but also because it promotes social cohesion and long-term investments in human capital \citep{benhabib2010economic, cox2018executive, suebvises2018social,ponzetto2018social}.

However, accountability systems can take many forms. One promising model for democratic oversight is _bottom-up accountability_, in which citizens receive information about the shortcomings of a given project to help them monitor and pressure underperforming public officials \citep{kosack2014does, molina2016community,raffler2018weakness}. Proponents suggest bottom-up accountability is effective because 1) constituents have first-hand information about the outcomes of local policies; 2) citizens have incentives to avoid corruption that directly affects them; 3) policy-makers are particularly sensitive to social punishment from their own communities \citep[570]{serra2011combining}. In a nutshell, bottom-up accountability offers a potential solution to the principal-agent dilemma in public service by aligning the interests of politicians and bureaucrats with those of the constituency they serve \citep[2]{raffler2018weakness}.

@@

[^transparencia]: Transparência Brasil is an non-governmental organisation whose mission is to 'promote transparency and social control of public power' in Brazil. It has been active since April 2000, is non-partisan and receives no public funding. More information is available at <http://transparencia.org.br> (access: July 2019).

@@

We used TDP to conduct three experimental interventions. We tested whether the app has a positive impact on school completion rates, school construction status and complaints to competent authorities. Contrary to our expectations, the TDP app has delivered mixed results. Two of the findings are positive and in line with our theoretical predictions. From August 2017 to February 2019, TDP increased the likelihood of school completion by 6.8 percent (SE = X.X, t-test = XXX). Additionally, we find that pressure over the bureaucracy increased the investments in the schools by 157.5 percent (SE = X.XX, t-test = XXX). In contrast, our results show the app had a null impact on school constructions statistics. Brazilian politicians did not respond to the citizen's requests placed via the application. Therefore, our findings raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. **Nota**: _Reescrever essa última frase e colocar mais explicações sobre possíveis mecanismos, etc_

@@

The _Tá de Pé_[^tdp] cell phone application is an initiative carried out by Transparência Brasil to foster bottom-up accountability in the Brazilian public sector. More specifically, the main goal of the TDP project is to improve the responsiveness in education infrastructure public expenditures. As we mentioned above, TDP incentivises citizens to provide up-to-date information about unfinished schools in their neighbourhoods, and these pictures are then assessed by engineers. In the case the construction is behind schedule, citizens are also provided with tools to contact their representatives quickly and anonymously. The app sends a notification to the mayor's office which has 15 days to answer. If they fail answer by the deadline, a notification goes to the federal government, making it harder to the municipality to access federal funds in the future. The motivation for the intervention is that information delivery causes individuals to closely monitor public works. This, in turn, results in better social outcomes as public agents become more responsible to community demands.

@@

We built the app in the first three months of 2017 and tested it in May 2017. The first stable Android version was deployed on Google Play on 14 August 2017. A version for iOS appeared two weeks later. At the same time, a massive Facebook campaign started in order to publicise the app. Facebook is the most widely used social media network in Brazil with around 72 million users in 2017 [@statista2019facebook]. The campaigns were concentrated in October 2017, investing around BRL 14.333,74 (US$ 4.479,29), and attracting 2028 new users in October 2017 only.

@@

\begin{figure}
     \begin{minipage}[t]{.3\textwidth}
       \centering
       \includegraphics[scale=0.25]{tdp01.jpg}
       \label{}
     \end{minipage}
     \hfill
     \begin{minipage}[t]{.3\textwidth}
       \centering
       \includegraphics[scale=0.25]{tdp02.jpg}
     \end{minipage}
     \hfill
     \begin{minipage}[t]{.3\textwidth}
       \centering
       \includegraphics[scale=0.25]{tdp03.jpg}
     \end{minipage}
\caption{The \textit{Tá de Pé} mobile phone application. The first image shows that a school construction is delayed by 9 months. The second image shows how citizens can add information to the photos they submit via the app. The last image presents a list of the closest school construction sites.} 
\label{fig:screenshots}
\end{figure}

>>>>

---
title:  |
        | The Pitfalls of Bottom-Up Accountability: Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Guilherme Fasolin, Stephen Herzog and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research and technical assistance. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "We study the effect of a mobile phone application designed to increase citizens' ability to monitor public works in Brazil. The app allowed individuals to find delayed school constructions in XXX Brazilian cities and send anonymous messages to politicians requestions information about construction delays and expected completion times. Our results show that the app has a null impact on the completion times subsequently reported by the municipalities. Additionally, we find that few politicians reacted to citizens' requests. These findings suggest it is difficult to motivate bottom-up accountability in new democracies, especially when politicians are unresponsive to non-electoral pressures. This paper has implications for the design of bottom-up accountability mechanisms."
abstractspacing: double
keywords: accountability, Brazil, field experiment, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 2cm
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

One of the key determinants of efficient public service provision is a robust accountability system [@besley2003incentives; @cameron2004public; @ferejohn1986incumbent; @o1998horizontal]. In its standard definition, accountability is understood as the process of holding authorities into account for their actions [@finer1941administrative; @mulgan2000accountability; @o1990bureaucratic]. Past studies show that accountability ensures that politicians act on behalf of voters [@freire2010ngp; @moncrieffe1998reconceptualizing], reduces the opportunities for rent-seeking and corruption [@deininger2005does; @wenar2006accountability] and increases the efficiency of the public sector [@adsera2003you; @bjorkman2009power]. Moreover, greater accountability is associated with higher levels of economic growth, not only because it limits state discretion in the economy, but also because it promotes social cohesion and long-term investments in human capital [@benhabib2010economic; @suebvises2018social; @ponzetto2018social].

However, accountability systems can take many forms. One promising model for democratic oversight is _bottom-up accountability_, in which citizens receive information about the shortcomings of a given project to help them monitor and pressure underperforming public officials [@kosack2014does; @molina2016community; @raffler2018weakness]. Proponents suggest bottom-up accountability is effective because 1) constituents have first-hand information about the outcomes of local policies; 2) citizens have incentives to avoid corruption that directly affects them; 3) policy-makers are particularly sensitive to social punishment from their own communities [@serra2011combining, 570]. In a nutshell, bottom-up accountability offers a potential solution to the principal-agent dilemma in public service by aligning the interests of politicians and bureaucrats with those of the constituency they serve [@barro1973control; @raffler2018weakness, 2].

@@

[^transparencia]: Transparência Brasil is an non-governmental organisation whose mission is to 'promote transparency and social control of public power' in Brazil. It has been active since April 2000, is non-partisan and receives no public funding. More information at <http://transparencia.org.br> (access: July 2019).

@@

We used TDP to conduct **two/three?** experimental interventions. We tested whether the app has a positive impact on **school completion rates, school construction status and complaints to competent authorities**. Contrary to our expectations, the TDP app has delivered mixed results. **One/Two** of the findings are positive and in line with our theoretical predictions. From August 2017 to February 2019, TDP increased the likelihood of school completion by 6.8 percent (SE = X.X, t-test = XXX). Additionally, we find that pressure over the bureaucracy increased the investments in the schools by 157.5 percent (SE = X.XX, t-test = XXX). In contrast, our results show the app had a null impact on school constructions statistics. Brazilian politicians did not respond to the citizen's requests placed via the application. Therefore, our findings raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. **Nota**: _Reescrever essa última frase e colocar mais explicações sobre possíveis mecanismos, etc_

@@

The _Tá de Pé_[^tdp] cell phone application is an initiative carried out by Transparência Brasil to foster bottom-up accountability in the Brazilian public sector. More specifically, the main goal of the TDP project is to improve the responsiveness in education infrastructure public expenditures. As we mentioned above, TDP incentivises citizens to provide up-to-date information about unfinished schools in their neighbourhoods, and these pictures are then assessed by engineers. In the case the construction is behind schedule, citizens are also provided with tools to contact their representatives quickly and anonymously. The app sends a notification to the mayor's office which has 15 days to answer. If they fail answer by the deadline, a notification goes to the federal government, making it harder to the municipality to access federal funds in the future. The motivation for the intervention is that information delivery causes individuals to closely monitor public works. This, in turn, results in better social outcomes as public agents become more responsive to community demands.

@@

We built the app in the first three months of 2017 and tested it in May 2017. The first stable Android version was deployed on Google Play on 14 August 2017. A version for iOS appeared six months later. A massive Facebook campaign started in order to publicise the app. Facebook is the most widely used social media network in Brazil with around 72 million users in 2017 [@statista2019facebook]. The campaigns were concentrated in October 2017, investing around BRL 14.333,74 (US$ 4.479,29), and attracting 2,028 new users in October 2017 only.

@@

\begin{figure}
     \begin{minipage}[t]{.3\textwidth}
       \centering
       \includegraphics[scale=0.25]{tdp03.jpg}
       \label{}
     \end{minipage}
     \hfill
     \begin{minipage}[t]{.3\textwidth}
       \centering
       \includegraphics[scale=0.25]{tdp01.jpg}
     \end{minipage}
     \hfill
     \begin{minipage}[t]{.3\textwidth}
       \centering
       \includegraphics[scale=0.25]{tdp02.jpg}
     \end{minipage}
\caption{The \textit{Tá de Pé} mobile phone application. The first image presents a list of the closest school construction sites. The second image shows that a school construction is delayed by 9 months. The last image shows how citizens can add information to the photos they submit via the app.} 
\label{fig:screenshots}
\end{figure}
